---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
require(logbin, quietly = TRUE)
```

# logbin

[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/logbin)](https://cran.r-project.org/package=logbin)

`logbin` provides methods for performing relative risk regression by fitting log-link GLMs and GAMs to binomial data. As well as providing a consistent interface to use the usual Fisher scoring algorithm (via `glm` or `glm2`) and an adaptive barrier approach (via `constrOptim`), it implements EM-type algorithms that have more stable convergence properties than other methods.

An example of periodic non-convergence using `glm` (run with `trace = TRUE` to see deviance at each iteration):

```{r, echo = TRUE, warning = FALSE}
require(glm2, quietly = TRUE)
data(heart)

start.p <- sum(heart$Deaths) / sum(heart$Patients)
t.glm <- system.time(
  fit.glm <- logbin(cbind(Deaths, Patients-Deaths) ~ factor(AgeGroup) + factor(Severity) + 
                      factor(Delay) + factor(Region), data = heart,
                    start = c(log(start.p), -rep(1e-4, 8)), method = "glm", maxit = 10000)
)
```

The combinatorial EM method (Marschner and Gillett, 2012) provides stable convergence:
```{r, echo = TRUE, message = FALSE}
t.cem <- system.time(fit.cem <- update(fit.glm, method = "cem"))
```

...but it can take a while. Using an overparameterised EM approach removes the need to run $3^4 = 81$ separate EM algorithms:
```{r, echo = TRUE, message = FALSE}
t.em <- system.time(fit.em <- update(fit.glm, method = "em"))
```

...while generic EM acceleration algorithms (from the `turboEM` package) can speed this up further still:
```{r, echo = TRUE, message = FALSE}
t.cem.acc <- system.time(fit.cem.acc <- update(fit.cem, accelerate = "squarem"))
t.em.acc <- system.time(fit.em.acc <- update(fit.em, accelerate = "squarem"))
```

Comparison of results:
```{r, echo = FALSE}
fit.list <- list(fit.glm, fit.cem, fit.em, fit.cem.acc, fit.em.acc)
time.list <- list(t.glm, t.cem, t.em, t.cem.acc, t.em.acc)
res <- data.frame(converged = sapply(fit.list, function(x) x$converged),
                  logLik = sapply(fit.list, logLik),
                  iterations = sapply(fit.list, function(x) x$iter[1]),
                  time = sapply(time.list, function(x) x[3]))
rownames(res) <- c("glm", "cem", "em", "cem.acc", "em.acc")
res
```

An adaptive barrier algorithm can also be applied using `method = "ab"`, with user-specified options via `control.method`: see `help(logbin)` for more details.

Semi-parametric regression using B-splines (Donoghoe and Marschner, 2015) can be incorporated by using the `logbin.smooth` function. See `example(logbin.smooth)` for a simple example.

## Installation

Get the released version from CRAN:

```{r, echo = TRUE, eval = FALSE}
install.packages("logbin")
```

Or the development version from github:

```{r, echo = TRUE, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mdonoghoe/logbin")
```

## References
* Donoghoe, M. W. and I. C. Marschner (2015). Flexible regression models for rate differences, risk differences and relative risks. _International Journal of Biostatistics_ __11__(1): 91--108.
* Donoghoe, M. W. and I. C. Marschner (2018). logbin: An R package for relative risk regression using the log-binomial model. _Journal of Statistical Software_ __86__(9): 1--22.
* Marschner, I. C. and A. C. Gillett (2012). Relative risk regression: reliable and flexible methods for log-binomial models. _Biostatistics_ __13__(1): 179--192.